---
title: "Modeling longitudinal traits with a random regression model"
subtitle: "Bridging the Gap: From Phenomics to Functional Genetics Workshop @HUJI"
author: "Gota Morota"
date: "April 2, 2019"
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: false
    code_folding: hide
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Background
We will model projected shoot area (PSA) collected over 20 time points from 89 temperate japonica rice accessions using a random regression model. 
```{r, eval=FALSE, echo=TRUE}
setwd("~/Dropbox/chikudaisei/teaching/2019/HUJI2019/lab2")
# install.packages("sommer")
# install.packages("orthopolynom")
# install.packages("graphics")
library("sommer")
library("orthopolynom")
library("graphics")

source("http://morotalab.org/Mrode2005/rr/stdtime.txt")
source("http://morotalab.org/Mrode2005/rr/legendre.txt")
```

## Read phenotype and genotype files
You can download the [G matrix](https://github.com/morota/HUJI2019/blob/gh-pages/lab2/G.txt) and [phenotype](https://github.com/morota/HUJI2019/blob/gh-pages/lab2/Pheno.txt) files [here](https://github.com/morota/HUJI2019/blob/gh-pages/lab2/).
```{r, eval=FALSE, echo=TRUE}
# G matrix
GRM <- as.matrix(read.table("G.txt", header = FALSE))
dim(GRM)

# Phenotype matrix
Pheno <-read.table("Pheno.txt", header = TRUE)        
head(Pheno)
# ID should be factor not numeric
Pheno$ID<- factor(Pheno$ID)                       
head(Pheno)
# Add column names
colnames(GRM) <- unique(Pheno$ID)
# Add row names
rownames(GRM) <- unique(Pheno$ID)                 
```

## Legendre polynomials
```{r, eval=FALSE, echo=TRUE}
Time= 1:20
# Order of fit
K <- 2
M <- stdtime(Time, K)
Lambda <- legendre(K, gengler = FALSE)
Phi <- M%*% t(Lambda)
matplot(Time, Phi, lwd=3, lty = 1, col = c("black","red","green"),type = "l",
       ylab = "Legendre polynomials",main="Quadratic Legendre polynomials")
```

## Fitting RRM
We will use the `mmer` function in the `sommer` R package. 
```{r, eval=FALSE, echo=TRUE}
RRM1 <-mmer(Y~1    
            ,random=~vs(leg(Time, 2), ID, Gu=GRM)
            ,rcov=~vs(ds(Time), units) 
            ,data=Pheno)
names(RRM1)
summary(RRM1)
````

## Predicting genetic values from random regression coefficients
We will predict genetic values from linear as well as quadratic Legendre polynomials.
```{r, eval=FALSE, echo=TRUE}
RnReg_Q <- rbind(unlist(RRM1$U$`leg0:ID`) # 3 x 89
               ,unlist(RRM1$U$`leg1:ID`) 
               ,unlist(RRM1$U$`leg2:ID`)
               )
colnames(RnReg_Q) <- unique(Pheno$ID)                
RnReg_L <- RnReg_Q[-3,]

Time <- unique(Pheno$Time)                        
Phi_Q <- leg(Time, 2)
Phi_L <- Phi_Q[, -3]
```

Prediction of genetic values from linear Legendre polynomials.
```{r, eval=FALSE, echo=TRUE}
GEBV_L <- t(apply(RnReg_L, 2, function (x) Phi_L %*% x)) 
GEBV2_L <- t(Phi_L %*% RnReg_L) # # alternative way
Pred_L <- cbind(Pheno, c(RRM1$Beta$Estimate + GEBV_L))
colnames(Pred_L)<- c("ID", "Pe", "Time","Observed","Predicted")
```

Prediction of genetic values from quadratic Legendre polynomials.
```{r, eval=FALSE, echo=TRUE}
GEBV_Q <- t(apply(RnReg_Q, 2, function (x) Phi_Q %*% x)) 
GEBV2_Q <- t(Phi_Q %*% RnReg_Q) # alternative way
Pred_Q <- cbind(Pheno, c(RRM1$Beta$Estimate + GEBV_Q))   
colnames(Pred_Q)<- c("ID", "Pe", "Time","Observed","Predicted")
```


## Comparison
Here, we will visualize observed vs. predicted longitudinal PSA for 12 accessions. 
```{r, eval=FALSE, echo=TRUE}
index <- c(1,9,10,15,31,32,36,51,52,56,62,63)      
```

Prediction based on linear Legendre polynomials.
```{r, eval=FALSE, echo=TRUE}
xyplot(Predicted+Observed~Time|ID,type ="l",  
       data = Pred_L[which(Pred_L$ID%in%index),], 
       pch=1, cex=1, auto.key=T, ylab="Phenotype")
```

Prediction based on quadratic Legendre polynomials.
```{r, eval=FALSE, echo=TRUE}
xyplot(Predicted+Observed~Time|ID,type ="l",
       data = Pred_Q[which(Pred_Q$ID%in%index),],  
       pch=1,cex=.8,auto.key=T, ylab="Phenotype")

```










