---
title: "Decoding mixed model equations"
subtitle: "Bridging the Gap: From Phenomics to Functional Genetics Workshop @HUJI"
author: "Gota Morota"
date: "April 1, 2019"
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: false
    code_folding: hide
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Background
```{r, eval=FALSE, echo=TRUE}
setwd("~/Dropbox/chikudaisei/teaching/2019/HUJI2019/lab2")
# install.packages("sommer")
# install.packages("orthopolynom")
install.packages("graphics")
# install.packages("data.table")
library("sommer")
library("orthopolynom")
library("graphics")
#library("data.table")

source("http://morotalab.org/Mrode2005/rr/stdtime.txt")
source("http://morotalab.org/Mrode2005/rr/legendre.txt")
```

## Read phenotype and genotype files
```{r, eval=FALSE, echo=TRUE}
# G matrix
GRM <- as.matrix(read.table("G.txt", header = FALSE))
dim(GRM)

# Phenotype matrix
Pheno <-read.table("Pheno.txt", header = TRUE)        
head(Pheno)
# ID should be factor not numeric
Pheno$ID<- factor(Pheno$ID)                       
head(Pheno)
# Add column names
colnames(GRM) <- unique(Pheno$ID)
# Add row names
rownames(GRM) <- unique(Pheno$ID)                 
```

## Legendre polynomials
```{r, eval=FALSE, echo=TRUE}
Time= 1:20
# Order of fit
K <- 2
M <- stdtime(Time, K)
Lambda <- legendre(K, gengler = FALSE)
Phi <- M%*% t(Lambda)
matplot(Time, Phi, lwd=3, lty = 1, col = c("black","red","green"),type = "l",
       ylab = "Legendre polynomials",main="Quadratic Legendre polynomials")
```



RRM.1 <-mmer(Y~1                                  # mmer function from sommer package && leg() from orthopolynom package, vs = variance structure, Gu = known covariance structures
            ,random=~vs(leg(Time, 2),ID,Gu=GRM)
            #,random=~vs(leg(Time, 2),ID,Gu=GRM) + vs(leg(Time, 2), Pe)
            ,rcov=~vs(units) # residuals, units = cov=~vs(ds(Time),units)
            #,rcov=~vs(ds(Time),units) # residuals, units = cov=~vs(ds(Time),units)
            ,data=Pheno)

names(RRM.1)
summary(RRM.1)

RnReg_Q <- rbind(unlist(RRM.1$U$`leg0:ID`),         # extract estimated random regression coeifficeints
               unlist(RRM.1$U$`leg1:ID`),         # Num. regression coefficients* Num. IDs. Dim is RnReg x nID = 3 x 89
               unlist(RRM.1$U$`leg2:ID`))



colnames(RnReg_Q)<- unique(Pheno$ID)                # Assign IDs to columns

Time <- unique(Pheno$Time)                         # Create a quadratic Legendre's
Phi_Q <- leg(Time, 2)
Phi_L <- Phi_Q[,-3]


RnReg_L <- RnReg_Q[-3,]
GEBV_L <- t(apply(RnReg_L, 2, function (x) Phi_L %*% x)) # estimating GEBVs for each IDs using RanReg: GEBV= Phi%*%RnReg
GEBV2_L <- t(Phi_L %*% RnReg_L)
Pred_L <- cbind(Pheno, c(RRM.1$Beta$Estimate + GEBV_L))   # merge orginal phenotype and predicted values and compare using lattice xyplot() function
colnames(Pred_L)<- c("ID", "Pe", "Time","Observed","Predicted")



IDs.f <- c(1,9,10,15,31,32,36,51,52,56,62,63)      # First 12 IDs


# Prediction using linear Leg
xyplot(Predicted+Observed~Time|ID,type ="l",       # xyplot() function from the lattice package
       data = Pred_L[which(Pred_L$ID%in%IDs.f),],  # To buil the appropriate plots
       pch=1, cex=1, auto.key=T, ylab="Phenotype")



GEBV_Q <- t(apply(RnReg_Q, 2, function (x) Phi_Q %*% x)) # estimating GEBVs for each IDs using RanReg: GEBV= Phi%*%RnReg
GEBV2_Q <- t(Phi_Q %*% RnReg_Q)
Pred_Q <- cbind(Pheno, c(RRM.1$Beta$Estimate + GEBV_Q))   # merge orginal phenotype and predicted values and compare using lattice xyplot() function
colnames(Pred_Q)<- c("ID", "Pe", "Time","Observed","Predicted")


xyplot(Predicted+Observed~Time|ID,type ="l",       # xyplot() function from the lattice package
       data = Pred_Q[which(Pred_Q$ID%in%IDs.f),],  # To buil the appropriate plots
       pch=1,cex=.8,auto.key=T, ylab="Phenotype")







RRM.2 <-mmer(Y~1                                  # mmer function from sommer package && leg() from orthopolynom package, vs = variance structure, Gu = known covariance structures
             ,random=~vs(leg(Time, 2),ID,Gu=GRM)
             #,random=~vs(leg(Time, 2),ID,Gu=GRM) + vs(leg(Time, 2), Pe)
             #,rcov=~vs(units) # residuals, units = cov=~vs(ds(Time),units)
             ,rcov=~vs(ds(Time),units) # residuals, units = cov=~vs(ds(Time),units)
             ,data=Pheno)








